Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-05-16 ms.Rnw
postpone examples
figures for examples

\citet{King2022} worked out formulas for the exact likelihood of a genealogy induced by an SIRS model.
The theory developed in this paper applies, but since there is only one deme in this model, this is a simple case.
Its state vector is $x=(S,I,R)$ and its KFE is
\begin{mathsize}{9pt}{10pt}
  \begin{equation*}
    \begin{split}
      \frac{\partial{v}}{\partial{t}}(t,S,I,R)=
      &\frac{\beta(t)\,(S+1)\,(I-1)}{N}\,v(t,S+1,I-1,R)
      -\frac{\beta(t)\,S\,I}{N}\,v(t,S,I,R)\\
      &+\gamma\,(I+1)\,v(t,S,I+1,R-1)
      -\gamma\,I\,v(t,S,I,R)\\
      &+\omega\,(R+1)\,v(t,S-1,I,R+1)
      -\omega\,R\,v(t,S,I,R).
    \end{split}
  \end{equation*}
\end{mathsize}%
Here $N=S+I+R$ is the host population size.
Note that we have here allowed for the possibility that the transmission rate, $\beta$, depends on time.

@@

A simple, yet interesting, model with more than one deme is the SEIRS model (\cref{fig:example_models}A).
The state space is $\Rp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
It has two demes ($\Demes=\Set{\lab{E},\lab{I}}$) and the KFE
\begin{mathsize}{9pt}{10pt}
  \begin{equation*}
    \begin{split}
      \frac{\partial{v}}{\partial{t}}(t,S,E,I,R)=
      &\frac{\beta(t)\,(S+1)\,I}{N}\,v(t,S+1,E-1,I,R)
      -\frac{\beta(t)\,S\,I}{N}\,v(t,S,E,I,R)\\
      &+\sigma\,(E+1)\,v(t,S,E+1,I-1,R)
      -\sigma\,E\,v(t,S,E,I,R)\\
      &+\gamma\,(I+1)\,v(t,S,E,I+1,R-1)
      -\gamma\,I\,v(t,S,E,I,R)\\
      &+\omega\,(R+1)\,v(t,S-1,E,I,R+1)
      -\omega\,R\,v(t,S,E,I,R),
    \end{split}
  \end{equation*}
\end{mathsize}%
where $N=S+E+I+R$ is the total population size.
The deme occupancy function in this case is $n(x)=(E,I)$.
Note that the terms associated with sampling cancel each other in the KFE, since, in this model, sampling has no effect on the state.

@@

A simple model for the competition of two strains for susceptible hosts is depicted in \cref{fig:example_models}B.
In this model, the state vector consists of seven numbers: $x=(S,E_1,E_2,I_1,I_2,R_1,R_2)$.
There are four demes ($\Demes=\Set{\lab{E}_1,\lab{E}_2,\lab{I}_1,\lab{I}_2}$) and the occupancy function is $n(x)=(E_1,E_2,I_1,I_2)$.
This model has KFE
\begin{mathsize}{9pt}{10pt}
  \begin{equation*}
    \begin{split}
      \frac{\partial{v}}{\partial{t}}=
      &\frac{\beta_1(t)\,I_1\,(S+1)}{N}\,v(t,S+1,E_1-1,E_2,I_1,I_2,R_1,R_2)
      -\frac{\beta_1(t)\,I_1\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
      &+\frac{\beta_2(t)\,I_2\,(S+1)}{N}\,v(t,S+1,E_1,E_2-1,I_1,I_2,R_1,R_2)
      -\frac{\beta_2(t)\,I_2\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
      &+\sigma_1\,(E_1+1)\,v(t,S,E_1+1,E_2,I_1-1,I_2,R_1,R_2)
      -\sigma_1\,E_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
      &+\sigma_2\,(E_2+1)\,v(t,S,E_1,E_2+1,I_1,I_2-1,R_1,R_2)
      -\sigma_2\,E_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
      &+\gamma_1\,(I_1+1)\,v(t,S,E_1,E_2,I_1+1,I_2,R_1-1,R_2)
      -\gamma_1\,I_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
      &+\gamma_2\,(I_2+1)\,v(t,S,E_1,E_2,I_1,I_2+1,R_1,R_2-1)
      -\gamma_2\,I_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
      &+\omega_1\,(R_1+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1+1,R_2)
      -\omega_1\,R_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
      &+\omega_2\,(R_2+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1,R_2+1)
      -\omega_2\,R_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    \end{split}
  \end{equation*}
\end{mathsize}%

@@

\Cref{fig:example_models}D depicts a model of superspreading.
There are three demes ($\Demes=\Set{\lab{E},\lab{I_L},\lab{I_H}}$).
The state vector is $x=(S,E,I_L,I_H,R)$ and the KFE is
\begin{mathsize}{9pt}{10pt}
  \begin{equation*}
    \begin{split}
      \frac{\partial{v}}{\partial{t}}=
      &\frac{\beta(t)\,(I_L+\theta\,I_H)\,(S+1)}{N}\,v(t,S+1,E-1,I_L,I_H,R)
      -\frac{\beta(t)\,(I_L+\theta\,I_H)\,S}{N}\,v(t,S,E,I_L,I_H,R)\\
      &\sigma\,(E+1)\,v(t,S,E+1,I_L-1,I_H,R)
      -\sigma\,E\,v(t,S,E,I_L,I_H,R)\\
      &\eps_{LH}\,(I_L+1)\,v(t,S,E,I_L+1,I_H-1,R)
      -\eps_{LH}\,I_L\,v(t,S,E,I_L,I_H,R)\\
      &\eps_{HL}\,(I_H+1)\,v(t,S,E,I_L-1,I_H+1,R)
      -\eps_{HL}\,I_H\,v(t,S,E,I_L,I_H,R)\\
      &\gamma\,(I_L+1)\,v(t,S,E,I_L+1,I_H,R-1)
      -\gamma\,I_L\,v(t,S,E,I_L,I_H,R)\\
      &\gamma\,(I_H+1)\,v(t,S,E,I_L,I_H+1,R-1)
      -\gamma\,I_H\,v(t,S,E,I_L,I_H,R)\\
      &\omega\,(R+1)\,v(t,S-1,E,I_L,I_H,R+1)
      -\omega\,R\,v(t,S,E,I_L,I_H,R).\\
    \end{split}
  \end{equation*}
\end{mathsize}%

@@

>>>>

\citet{King2022} worked out formulas for the exact likelihood of a genealogy induced by an SIRS model.
The theory developed in this paper applies, but since there is only one deme in this model, this is a simple case.
Its state vector is $x=(S,I,R)$ and its KFE is
\begin{mathsize}{9pt}{10pt}
  \begin{equation*}
    \begin{split}
      \frac{\partial{v}}{\partial{t}}(S,I,R)=
      &\frac{\beta\,(S+1)\,(I-1)}{N}\,v(t,S+1,I-1,R)
      -\frac{\beta\,S\,I}{N}\,v(t,S,I,R)\\
      &+\gamma\,(I+1)\,v(t,S,I+1,R-1)
      -\gamma\,I\,v(t,S,I,R)\\
      &+\omega\,(R+1)\,v(t,S-1,I,R+1)
      -\omega\,R\,v(t,S,I,R).
    \end{split}
  \end{equation*}
\end{mathsize}%
Here $N=S+I+R$ is the host population size.
Note that, though the theory allows for time-dependent event rates, this example is time-homogenous.
This model has one deme, $\Demes=\Set{\lab{I}}$, with occupancy function $n(x)=I$.
There are four kinds of jumps:
transmission, recovery, waning of immunity, and sampling.
Accordingly, the marks are $\Jumps=\Set{\lab{Trans},\lab{Recov},\lab{Wane},\lab{Sample}}$.
\Cref{tab:sirs_model_elements} gives $\alpha_u$, $r_u$, and the jump-type for each of these marks.

\begin{table}
  \caption{
    \label{tab:sirs_model_elements}
    Elements of the SIRS model pertinent to the genealogy process.
  }
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{cclc}
    \hline\hline
    $u$ & $\alpha_u$ & Event type & $r_u$ \\
    \hline
    $\lab{Trans}$  & ${\beta\,S\,I}/{N}$  & pure birth  & 2 \\
    $\lab{Recov}$  & $\gamma\,I$          & pure death  & 0 \\
    $\lab{Wane}$   & $\omega\,R$          & neutral     & 0 \\
    $\lab{Sample}$ & $\psi\,I$            & pure sample & 1 \\
    \hline\hline
  \end{tabular}
\end{table}

Given an obscured genealogy $Z$, let $\event{Z}=B\cup{S_0}\cup{S_1}$, where $B$ is the set of branch-times, and $S_0$, $S_1$ are the sets of sample-times with saturations $0$ and $1$, respectively.
Since there is only one deme, paintings of $Z$ can differ only in number and position of inline, internal nodes along branches.
Each of these can only correspond to $u=\lab{Trans}$ with $s=1$.
For $t\notin{\event{Z}}$, we can take the importance sampling distribution to be
\begin{equation*}
  \pi_u=\begin{cases}
  c,                & u=\lab{Trans}, s=0, t\notin{\event{Z}},\\
  \frac{1-c}{\ell}, & u=\lab{Trans}, s=1, t\notin{\event{Z}},\\
  0,                & \text{otherwise}.
  \end{cases}
\end{equation*}
Here $c$ is an arbitrary probability that does not affect the computation.
The relevant binomial ratios are
\begin{equation*}
  \BinRatio{n}{\ell}{r}{s}=
  \begin{cases}
    \frac{(I-\ell)\,(I-\ell-1)}{I\,(I-1)}, & u=\lab{Trans}, s=0, \\[1ex]
    \frac{2\,(I-\ell)}{I\,(I-1)}, & u=\lab{Trans}, s=1, \\[1ex]
    \frac{2}{I\,(I-1)}, & u=\lab{Trans}, s=2, \\[1ex]
    %% 1, & u=\lab{Recov}, I\ge{\ell}, \\[1ex]
    %% 0, & u=\lab{Recov}, I=\ell, \\[1ex]
    %% 1, & u=\lab{Wane}, \\[1ex]
    \frac{I-\ell}{I}, & u=\lab{Sample}, s=0, \\[1ex]
    \frac{1}{I}, & u=\lab{Sample}, s=1. \\[1ex]
  \end{cases}
\end{equation*}

This leads, for $t\notin{\event{Z}}$, to the following regular part of the filter equation:
\begin{mathsize}{9pt}{10pt}
  \begin{equation*}
    \begin{split}
      \frac{\partial{w}}{\partial{t}}=
      &\frac{\beta\,(S+1)\,(I-1)}{N}\,\left(1-\frac{\tbinom{\ell(t)}{2}}{\tbinom{I}{2}}\right)\,w(t,S+1,I-1,R)
      -\frac{\beta\,S\,I}{N}\,w(t,S,I,R)\\
      &+\gamma\,(I+1)\,w(t,S,I+1,R-1)
      -\gamma\,I\,w(t,S,I,R)\\
      &+\omega\,(R+1)\,w(t,S-1,I,R+1)
      -\omega\,R\,w(t,S,I,R)\\
      &-\psi\,I\,w(t,S,I,R).
    \end{split}
  \end{equation*}
\end{mathsize}%
Here, we have summed over the various paintings for $u=\lab{Trans}$, $s<2$.
Note the presence of the decay term proportional to $\psi$.
At event-times $t\in\event{Z}$, the singular part of the filter equation reads
\begin{mathsize}{9pt}{10pt}
  \begin{equation*}
    w(t,S,I,R)=
    \begin{cases}
      \wt(t,S+1,I-1,R)\,\frac{2\,\beta\,(S+1)}{N\,I}, & t\in{B}, \\[1ex]
      \wt(t,S,I,R)\,\psi\,(I-\ell(t)), & t\in{S_0}, \\[1ex]
      \wt(t,S,I,R)\,\psi, & t\in{S_1}. \\[1ex]
    \end{cases}
  \end{equation*}
\end{mathsize}%
Finally, note that $w(t,S,I,R)=0$ for all $I<\ell(t)$.

<<sirs3>>=
bake(
  file="sirs3a.rds",
  seed=328168304L,
  {
    library(phylopomp)
    runSIRS(
      Beta=4,gamma=2,psi=1,omega=1,
      S0=97,I0=3,R0=0,t0=0,time=40
    )
  }
) -> x

bake(
  file="sirs3b.rds",
  seed=621400057L,
  {
    expand_grid(
      Beta=4,
      ##      gamma=2,
      gamma=seq(1.7,2.3,by=0.02),
      psi=1,
      omega=1,
      ##      omega=seq(0.5,2,by=0.05),
      S0=97,I0=3,R0=0,
      t0=0,
      rep=1:16,
      Np=5000,
      ) -> params

    library(iterators)
    library(doFuture)
    plan(multicore)

    foreach (
      p=iter(params,"row"),
      .combine=bind_rows,
      .options.future=list(seed=TRUE)
    ) %dofuture% {
      library(pomp)
      library(phylopomp)
      p |>
        with({
          x |>
            sirs_pomp(
              Beta=Beta,gamma=gamma,psi=psi,omega=omega,
              S0=S0,I0=I0,R0=R0,t0=0
            ) |>
            pfilter(Np=Np) |>
            logLik()
        }) -> ll
      bind_cols(p,logLik=ll)
    } -> params
  }
) -> params
@

\begin{figure}
  \begin{center}
    <<sirs3_plot,dependson="sirs1",fig.dim=c(8,2.8),out.width="100%">>=
    plot_grid(
      A=x |>
        plot(points=FALSE,palette=c("#000000"))+
        labs(x="time"),
      B=params |>
        group_by(Beta,gamma,psi,omega,S0,I0,R0,Np) |>
        reframe(
          name=c("logLik","logLik_se"),
          value=logmeanexp(logLik,se=TRUE)
        ) |>
        ungroup() |>
        pivot_wider() |>
        mutate(
          y=logLik,
          ymax=logLik+2*logLik_se,
          ymin=logLik-2*logLik_se
        ) |>
        filter(round(Beta,2)==4) |>
        filter(logLik>max(logLik)-16) |>
        ggplot(
          aes(
            x=gamma,
            group=Np,
            y=y,
            ymin=ymin,
            ymax=ymax
          )
        )+
        geom_errorbar(position="dodge")+
        geom_vline(xintercept=2,color="red")+
        labs(
          color=character(0),
          y="log likelihood",
          x=expression(gamma)
        )+
        theme_classic()+
        theme(
          legend.position=c(0.5,0.27)
        ),
      labels="AUTO",
      nrow=1,
      rel_widths=c(1,1)
    )
    @
  \end{center}
\end{figure}

@@

A simple, yet interesting, model with more than one deme is the SEIRS model (\cref{fig:example_models}A).
The state space is $\Rp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
The KFE for the population process is
\begin{mathsize}{9pt}{10pt}
  \begin{equation*}
    \begin{split}
      \frac{\partial{v}}{\partial{t}}(t,S,E,I,R)=
      &\frac{\beta(t)\,(S+1)\,I}{N}\,v(t,S+1,E-1,I,R)
      -\frac{\beta(t)\,S\,I}{N}\,v(t,S,E,I,R)\\
      &+\sigma\,(E+1)\,v(t,S,E+1,I-1,R)
      -\sigma\,E\,v(t,S,E,I,R)\\
      &+\gamma\,(I+1)\,v(t,S,E,I+1,R-1)
      -\gamma\,I\,v(t,S,E,I,R)\\
      &+\omega\,(R+1)\,v(t,S-1,E,I,R+1)
      -\omega\,R\,v(t,S,E,I,R),
    \end{split}
  \end{equation*}
\end{mathsize}%
where $N=S+E+I+R$ is the total population size.
This model has two demes: $\Demes=\Set{\lab{E},\lab{I}}$ and four
The deme occupancy function in this case is $n(x)=(E,I)$.
Note that the terms associated with sampling cancel each other in the KFE, since, in this model, sampling has no effect on the state.

@@

A simple model for the competition of two strains for susceptible hosts is depicted in \cref{fig:example_models}B.
This example will be completed in a forthcoming draft.

@@

\Cref{fig:example_models}D depicts a model of superspreading.
This example will be completed in a forthcoming draft.

@@

<<lbdp3>>=
true.params <- data.frame(lambda=1.2,mu=0.8,psi=1,n0=5)

bake(
  file="lbdp3a.rds",
  seed=505641575,  
  true.params |>
    with(
      runLBDP(lambda=lambda,mu=mu,psi=psi,n0=n0,time=5)
    ) -> x
) -> x

bake(
  file="lbdp3b.rds",
  dependson=x,
  seed=715604304,
  {
    expand_grid(
      rep=1:10,
      lambda=1.2,
      mu=seq(0.3,2,by=0.1),
      ## mu=0.8,
      ## psi=seq(0.2,2.5,by=0.1),
      psi=1,
      time=3,
      n0=5,
      Np=c(1000,2000,4000,8000,16000,32000,64000,128000)
    ) -> params

    library(iterators)
    library(doFuture)
    plan(multicore)

    foreach (
      p=iter(params,"row"),
      .combine=bind_rows,
      .options.future=list(seed=TRUE)
    ) %dofuture% {
      p |>
        with(
          x |>
            lbdp_exact(lambda=lambda,mu=mu,psi=psi,n0=n0)
        ) -> ll1
      p |>
        with(
          x |>
            lbdp_pomp(lambda=lambda,mu=mu,psi=psi,n0=n0) |>
            pfilter(Np=Np) |>
            logLik()
        ) -> ll2
      bind_cols(p,exact=ll1,pf=ll2)
    }-> params
  }
) -> params

params |>
  mutate(
    diff=(pf-exact)/exact
  ) |>
  group_by(Np) |>
  summarize(
    rmse=sqrt(mean(diff*diff)),
    sd=sd(diff),
    bias=abs(mean(pf-exact))
  ) |>
  ungroup() -> stats
@

\begin{figure}
  \begin{center}
    <<lbdp3_plot,fig.dim=c(8,5),out.width="100%">>=
    pal <- c(viridis_pal(option="H",begin=0.1,end=0.8)(8),"#000000")
    names(pal) <- c("1k","2k","4k","8k","16k","32k","64k","128k","exact")

    plot_grid(
      params |>
        pivot_longer(c(exact,pf)) |>
        unite(name,name,Np) |>
        mutate(
          name=if_else(grepl("exact",name),"exact",name),
          name=gsub("pf_","",name),
          name=gsub("000","k",name),
          name=ordered(name,levels=names(pal))
        ) |>
        group_by(lambda,mu,psi,time,n0,name) |>
        reframe(
          type=c("logLik","logLik_se"),
          value=logmeanexp(value,se=TRUE)
        ) |>
        ungroup() |>
        pivot_wider(names_from=type) |>
        mutate(
          y=logLik,
          ymax=logLik+2*logLik_se,
          ymin=logLik-2*logLik_se
        ) |>
        filter(logLik>max(logLik)-16) |>
        ggplot(aes(x=mu,group=name,color=name,
          y=y,ymin=ymin,ymax=ymax))+
        geom_errorbar(
          position="dodge"
        )+
        geom_vline(xintercept=true.params$mu,linetype=2)+
        geom_hline(
          yintercept=max(params$exact)-
            c(0,0.5*qchisq(p=0.95,df=1)),
          linetype=2
        )+
        scale_color_manual(values=pal)+
        labs(
          color="effort",
          y="log likelihood",
          x=expression(mu)
        )+
        theme(
          legend.position="inside",
          legend.position.inside=c(0.5,0.4),
          legend.background=element_rect(fill="white")
        ),
      plot_grid(
        ##        x |> plot(points=TRUE),
        stats |>
          ggplot(aes(x=Np,y=rmse))+
          geom_smooth(formula=y~x,method="lm")+
          geom_point()+
          scale_x_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
          scale_y_log10()+
          coord_fixed(ratio=1)+
          labs(x="effort",y="RMSE"),
        stats |>
          ggplot(aes(x=Np,y=sd))+
          geom_smooth(formula=y~x,method="lm")+
          geom_point()+
          scale_x_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
          scale_y_log10()+
          coord_fixed(ratio=1)+
          labs(x="effort",y="SD"),
        stats |>
          ggplot(aes(x=Np,y=bias))+
          geom_smooth(formula=y~x,method="lm")+
          geom_point()+
          scale_x_log10(labels=\(x)aakmisc::scinot(x,simplify=TRUE))+
          scale_y_log10()+
          coord_fixed(ratio=1)+
          labs(x="effort",y="bias"),
        ncol=1,
        align="h",
        axis="btlr",
        rel_heights=c(1,1,1)
      ),
      nrow=1,
      rel_widths=c(2,1)
    )
    @
  \end{center}
\end{figure}
